%!TEX root = da-screen.tex

This book is an introduction to the theory of distributed algorithms. The topics covered include:
\begin{itemize}
    \item \textbf{Models of computing:} precisely what is a distributed algorithm, and what do we mean when we say that a distributed algorithm solves a certain computational problem?
    \item \textbf{Algorithm design and analysis:} which computational problems can be solved with distributed algorithms, which problems can be solved efficiently, and how to do it?
    \item \textbf{Computability and computational complexity:} which computational problems \emph{cannot} be solved at all with distributed algorithms, which problems cannot be solved efficiently, and why is this the case?
\end{itemize}
No prior knowledge of distributed systems is needed. A basic knowledge of discrete mathematics and graph theory is assumed, as well as familiarity with the basic concepts from undergraduate-level courses on models on computation, computational complexity, and algorithms and data structures.


\usection{About the Course}

This textbook was written to support the lecture course \emph{CS-E4510 Distributed Algorithms} at Aalto University. The course is worth 5 ETCS credits. There are 12 weeks of lectures: each week there is one 2-hour lecture and one 2-hour exercise session. Each week we will cover one chapter of this book. In each chapter there are at least 5 exercises, and the students are expected to solve at least 3 of them.

Prior versions of this textbook have been used in the lecture course of \emph{Deterministic Distributed Algorithms}, which I lectured at the University of Helsinki in 2010, 2012, and 2014.


\usection{Acknowledgements}

Many thanks to Jaakko Alasaari, Alkida Balliu, Mika G\"o\"os, Juho Hirvonen, Joel Kaasinen, Mirco Kroon, Tee\-mu Kuu\-sisto, Dang Lam, Tuo\-mo Lem\-pi\"a\-inen, Christoph Lenzen, Abdulmelik Mohammed, Christopher Purcell, Joel Rybicki, Stefan Schmid, Roelant Stegmann, Przemys\l{}aw Uzna\'nski, and Jussi V\"ai\-s\"a\-nen for feedback, discussions, comments, and for helping me with the arrangements of this course. This work was supported in part by the Academy of Finland, Grant 252018. For updates and additional material, see
\begin{center}
    \url{https://jukkasuomela.fi/da/}
\end{center}


\usection{License}

\begin{samepage}
This work is licensed under the \emph{Creative Commons Attribution--ShareAlike 3.0 Unported License}. To view a copy of this license, visit
\begin{center}
    \url{http://creativecommons.org/licenses/by-sa/3.0/}
\end{center}
\end{samepage}
