%!TEX root = da-dev.tex

\usection{Notation}

{\raggedright
\begin{notation}
    \item[$|X|$] the number of elements in set $X$
    \item[$f^{-1}(y)$] preimage of $y$, i.e., $f^{-1}(y) = \Set{ x : f(x) = y }$
    \item[${}^i 2$] power tower, ${}^i 2 = 2^{2^{\cdot^{\cdot^2}}}$ with $i$ twos
    \item[$\log^* n$] iterated logarithm, $\log^*({}^i 2) = i$
    \item[$\deg_G(v)$] degree of node $v$ in graph $G$
    \item[$\dist_G(u,v)$] distance between nodes $u$ and $v$ in $G$
    \item[$\ball_G(v,r)$] nodes that are within distance $r$ from $v$ in $G$
    \item[$\diam(G)$] diameter of graph $G$
    \item[$\NN$] the set of natural numbers, $\{0,1,2,\dotsc\}$
    \item[$\NNpos$] the set of positive integers, $\{1,2,\dotsc\}$
    \item[$\RR$] the set of real numbers
    \item[{$[a,b]$}] set $\{x \in \RR : a \le x \le b\}$
    \item[$Y^{(k)}$] the collection of all $k$-subsets of $Y$
    \item[$R_c(n;k)$] Ramsey numbers
\end{notation}}

\usection{Symbols}

These conventions are usually followed in the choice of symbols.

{\raggedright
\begin{notation}
    \item[$\alpha$] approximation factor
    \item[$\chi$] range of unique identifiers, $\chi = |V|^c$
    \item[$\phi$] covering map, $\phi \colon V \to V'$
    \item[$\psi$] local isomorphism, $\psi \colon \ball_G(v,r) \to \ball_H(u,r)$
    \item[$\Delta$] maximum degree; an upper bound of the maximum degree
    \item[$\Pi$] graph problem
    \item[$\calF$] graph family
    \item[$\calS$] set of feasible solutions
    \item[$\Id$] unique identifiers
    \item[$A$] distributed algorithm
    \item[$C$] vertex cover $C \subseteq V$, edge cover $C \subseteq E$
    \item[$D$] dominating set $D \subseteq V$, edge dominating set $D \subseteq E$
    \item[$E$] set of edges
    \item[$G$, $H$] graph, $G = (V,E)$
    \item[$I$] independent set $I \subseteq V$
    \item[$M$] matching $M \subseteq E$
    \item[$N$] port-numbered network, $N = (V,P,p)$
    \item[$P$] set of ports
    \item[$T$] running time (number of rounds)
    \item[$U$] subset of nodes
    \item[$V$] set of nodes
    \item[$c,C,d$] natural number
    \item[$e$] edge, element of $E$
    \item[$f,g,h$] function
    \item[$i,j,k,\ell$] natural number
    \item[$m_t$] message
    \item[$m$] number of edges, $m = |E|$
    \item[$n$] number of nodes, $n = |V|$
    \item[$p$] connection function, involution $p \colon P \to P$
    \item[$r$] natural number
    \item[$s,t,u,v$] node, element of $V$
    \item[$t$] time step (round), $t = 0, 1, \dotsc, T$
    \item[$w$] walk
    \item[$x_t$] state
\end{notation}}

\usection{Models of Computing}

{\raggedright
\begin{notation}
    \item[$\PN$] port-numbering model, Chapter~\ref{ch:pn}.
    \item[$\LOCAL$] networks with unique identifiers, Chapter~\ref{ch:local}.
    \item[$\CONGEST$] bandwidth-limited networks, Chapter~\ref{ch:congest}.
\end{notation}}

\usection{Algorithms}

\begin{algorithms}
    \item[$\algo{P3C}$] Deterministic algorithm for $3$-colouring a path. Runs in $O(n)$ rounds in the $\LOCAL$ model. Sections \ref{sec:algo-p3c} and \ref{sec:algo-p3c-formal}.
    \item[$\algo{P3CBit}$] Deterministic algorithm for $3$-colouring a directed path. Runs in $O(\log^* n)$ rounds in the $\LOCAL$ model. Section~\ref{sec:algo-p3cbit}.
    \item[$\algo{P3CRand}$] Randomised algorithm for $3$-colouring a path. Runs in $O(\log n)$ rounds with high probability. Section~\ref{sec:algo-p3crand}.
    \item[$\algo{P2C}$] Deterministic algorithm for $2$-colouring a path. Runs in $O(n)$ rounds in the $\LOCAL$ model. Section~\ref{ssec:algo-p2c}.
    \item[$\algo{BMM}$] Deterministic algorithm for finding a maximal matching in $2$-coloured graphs. Runs in $O(\Delta)$ rounds in the $\PN$ model. Section~\ref{sec:bmm}.
    \item[$\algo{VC3}$] Deterministic algorithm for finding a \Apx{3} of minimum vertex cover. Runs in $O(\Delta)$ rounds in the $\PN$ model. Section~\ref{sec:vc3}.
    \item[$\algo{Gather}$] Deterministic algorithm for gathering information on the communication network. Runs in $O(\diam(G))$ rounds in the $\LOCAL$ model. Section~\ref{sec:gather}.
    \item[$\algo{Greedy}$] Greedy colour reduction. Section~\ref{sec:greedy}.
    \item[$\algo{DPGreedy}$] Greedy colour reduction in directed pseudoforests. Section~\ref{sec:dpgreedy}.
    \item[$\algo{DPBit}$] Fast colour reduction in directed pseudoforests. Section~\ref{sec:dpbit}.
    \item[$\algo{DPSet}$] Fast colour reduction in directed pseudoforests. Exercise~\ref{ex:dpset}.
    \item[$\algo{DP3C}$] Deterministic algorithm for $3$-colouring a directed pseudoforest. Runs in $O(\log^* n)$ rounds in the $\LOCAL$ model. Section~\ref{sec:dp3c}.
    \item[$\algo{BDColour}$] Deterministic algorithm for $(\Delta+1)$-colouring graphs of maximum degree $\Delta$. Runs in $O(\Delta^2 + \log^* n)$ rounds in the $\LOCAL$ model. Section~\ref{sec:bdcolour}.
\end{algorithms}
